---
title: "nFK Reference Cases"
output:
  html_document:
    theme: cosmo
    df_print: paged
---

```{r setup, include=FALSE}
suppressMessages({ library(devtools); load_all(quiet = TRUE) })
library(knitr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{css, echo=FALSE}
table.nfk-ref { border-collapse: collapse; width: 100%; margin-bottom: 1.5em; font-size: 0.95em; }
table.nfk-ref td { border: 1px solid #ddd; padding: 9px 12px; vertical-align: top; }
table.nfk-ref th { border: 1px solid #ddd; padding: 7px 12px; background: #f5f5f5; text-align: left; }
table.nfk-ref pre { margin: 0; font-size: 0.82em; line-height: 1.22; }
table.nfk-ref td:nth-child(n+3),
table.nfk-ref th:nth-child(n+3) {
  text-align: right;
  font-variant-numeric: tabular-nums;
  white-space: nowrap;
  padding-left: 8px;
  padding-right: 10px;
}
table.nfk-ref td:nth-child(3) { font-weight: 600; color: #2a4d8f; }
table.nfk-ref th:nth-child(3) { color: #2a4d8f; }
.nfk-desc { color: #333; }
.nfk-ex { font-weight: 600; color: #555; margin-right: 0.25em; }
```

`nFK` is the sum of per-component **&omega;<sub>max</sub>** values (Bentler 1968; Lord 1958): for each principal component of the input matrix, &omega;<sub>max</sub> is the maximum reliability achievable by the optimal linear composite of items measuring that component. Reading the totals below, each `nFK` value is "how many factors' worth of reliable measurement the item set provides," with each factor weighted by how well items recover it.

## How nFK is computed

Let $R$ be a $p \times p$ input matrix with reliability or communality on the diagonal and inter-item correlations off-diagonal. The construction proceeds in five steps.

**1. Eigendecomposition and loadings.** Eigendecompose $R = V \Lambda V'$ with $\Lambda = \mathrm{diag}(\lambda_1, \ldots, \lambda_p)$, and form the loading matrix

$$L \;=\; V \,\mathrm{diag}\!\bigl(\sqrt{\lambda_1^{+}}, \ldots, \sqrt{\lambda_p^{+}}\bigr), \qquad \lambda_k^{+} = \max(\lambda_k, 0).$$

Column $L_{\cdot k}$ is the loading of items on the $k$-th principal component. (When `rotate = "varimax"`, an orthogonal rotation is applied to $L$ at this point. Total nFK is invariant under orthogonal rotation; the per-component values redistribute.)

**2. Joint item+component matrix.** Stack items and components into a $2p \times 2p$ block matrix and force the full diagonal to 1:

$$B \;=\; \begin{bmatrix} R & L \\ L' & \mathbf{0} \end{bmatrix}, \qquad \mathrm{diag}(B) \,:=\, 1.$$

The diagonal swap standardizes items (top-left block) and components (bottom-right block) to unit variance; the reliability information that previously lived on the diagonal of $R$ is preserved in $L$, since the eigenvalues of $R$ shrink with the diagonal.

**3. Smooth to the nearest correlation matrix.** $B$ is rank-deficient by construction (components are exact linear combinations of items), so apply Higham's (2002) algorithm to project to the nearest valid correlation matrix:

$$\tilde{B} \;=\; \arg\!\min_{X \,\succeq\, 0,\;\mathrm{diag}(X)\,=\,1} \;\|X - B\|_{F}.$$

For well-behaved inputs the perturbation is negligible; the smoothing matters mainly when $R$ is rank-deficient or pseudo-non-PSD (e.g., reliability lower than off-diagonal correlation), where it is what guarantees $\tilde{R}_{xx}$ is invertible.

**4. Per-component squared multiple correlation.** Partition $\tilde{B}$ back into its top-left item block $\tilde{R}_{xx}$ and top-right item-component block $\tilde{L}$, then for each component $k$,

$$R^{2}_{k} \;=\; \tilde{L}_{\cdot k}' \, \tilde{R}_{xx}^{-1} \, \tilde{L}_{\cdot k}.$$

This is the squared multiple correlation between items and component $k$, which equals $\omega_{\max, k}$ (Bentler 1968; Lord 1958): the maximum reliability achievable by the optimal linear composite of items measuring that component.

**5. Sum.**

$$\mathrm{nFK} \;=\; \sum_{k=1}^{p} R^{2}_{k} \;=\; \sum_{k=1}^{p} \omega_{\max, k}.$$

**Reading the statistic.** Each $R^{2}_{k} \in [0, 1]$ is the reliability of the best item-weighted composite for that dimension; $\mathrm{nFK}$ is "how many factors' worth of reliable measurement" the item set provides. When items measuring a factor are tau-equivalent (uniform loadings, uniform within-cluster correlation), $\omega_{\max}$ collapses to Cronbach's $\alpha$, and for symmetric scales further to Spearman-Brown. When $R$ is rank-deficient, components with zero loadings contribute zero, so $\mathrm{rank}(R)$ is an upper bound on $\mathrm{nFK}$. When $\mathrm{diag}(R) = 1$ throughout (no measurement error encoded), every component is exactly recoverable from items and $\mathrm{nFK} = p$ regardless of off-diagonal magnitude &mdash; the off-diagonal correlation is read as true-score association, not as a measurement-information loss. To register inter-item correlation as measurement error, the reduction has to live on the diagonal as reliability or communality below 1.

## Other statistics for indexing the effective number of factors or variables in a set

The seven comparison columns to the right of `nFK` come from `poolr::meff()` (Nyholt 2004; Li &amp; Ji 2005; Gao et al. 2008; Galwey 2009; Chen et al. 2012), with two adjustments to poolr's defaults: the final `floor()` that poolr applies for Bonferroni correction is removed (so continuous gradations remain visible), and the two scale-invariant methods (Gao, Galwey) are reported a second time on the matrix as supplied (without forcing diag = 1). Each method reduces to a function of either eigenvalues or matrix entries.

Throughout, $\lambda_1, \ldots, \lambda_p$ are the eigenvalues of the diag-1 correlation matrix and $\lambda_{(j)}$ is the $j$-th largest. For the starred variants, $\mu_k$ replaces $\lambda_k$ &mdash; eigenvalues of the input matrix as supplied (diagonal preserved as reliability or communality). $\mathbb{I}(\cdot)$ is the indicator function and $\lfloor \cdot \rfloor$ is the floor.

**Nyholt (2004).** Moment-based; returns $p$ for equal eigenvalues (no shared variance) and 1 for a rank-1 spectrum. Bakes in the correlation-matrix assumption $\mathrm{mean}(\lambda) = 1$.

$$M_{\text{eff}} \;=\; 1 + (p - 1)\!\left(1 - \frac{\mathrm{Var}(\lambda)}{p}\right)$$

**Li &amp; Ji (2005).** Each eigenvalue $\geq 1$ counts as a full factor (Kaiser's rule); fractional parts contribute partially.

$$M_{\text{eff}} \;=\; \sum_{k=1}^{p}\!\left[\,\mathbb{I}(|\lambda_k| \geq 1) + \bigl(|\lambda_k| - \lfloor |\lambda_k| \rfloor\bigr)\right]$$

**Gao et al. (2008).** Smallest number of leading components needed to capture a fraction $C$ of the eigenvalue sum (poolr default $C = 0.995$). Intrinsically integer.

$$M_{\text{eff}} \;=\; \min\!\left\{k \;:\; \frac{\sum_{j=1}^{k} \lambda_{(j)}}{\sum_{j=1}^{p} \lambda_j} > C\right\}$$

**Galwey (2009).** The participation ratio of the eigenvalue spectrum under a square-root transform of the non-negative eigenvalues. Returns $p$ for an equal spectrum, 1 for rank 1.

$$M_{\text{eff}} \;=\; \frac{\bigl(\sum_{k=1}^{p} \sqrt{\lambda_k^{+}}\,\bigr)^2}{\sum_{k=1}^{p} \lambda_k^{+}}, \qquad \lambda_k^{+} = \max(\lambda_k, 0)$$

**Chen et al. (2012).** Acts on the matrix entries directly via row $\ell_C$ norms (poolr default $C = 7$). Assumes $|R_{ij}| \leq 1$ throughout, so it cannot be transferred to a non-correlation matrix.

$$M_{\text{eff}} \;=\; \sum_{i=1}^{p} \frac{1}{\sum_{j=1}^{p} |R_{ij}|^{C}}$$

**Gao\* and Galwey\* (as supplied).** Identical formulas to Gao and Galwey, evaluated on $\mu_k$ rather than $\lambda_k$. Only these two transfer cleanly: their formulas are scale-invariant in the eigenvalue spectrum, so they remain meaningful when the diagonal carries reliability rather than 1. Nyholt's $\mathrm{mean}(\lambda) = 1$, Li &amp; Ji's threshold of 1, and Chen's $|R_{ij}| \leq 1$ are all violated on the as-supplied matrix.

Each row of the table shows a 6&times;6 input matrix (diagonal = reliability, off-diagonal = inter-item correlation) alongside `nFK` and the seven variants above. The five `meff` columns are computed on a copy with diag forced to 1; `nFK`, `gao*`, and `galwey*` consume the matrix as supplied. Rows that share an off-diagonal pattern but differ in diagonal reliability (e.g., #3 vs #6 vs #8, or #5 vs #7) collapse to identical values across the five `meff` columns, while `nFK`, `gao*`, and `galwey*` separate them.

```{r references}
varnames <- c("A", "B", "C", "D", "E", "F")

mk <- function(off, dg = 1) {
  m <- matrix(off, 6, 6); diag(m) <- dg; m
}

ref <- list(
  list(
    ex          = "#1",
    diag        = 1.0, r = 0.0,
    description = "Identity matrix. Six perfectly reliable, uncorrelated items &mdash; six 1-item composites each at reliability 1. nFK = p = 6 is the upper bound.",
    m           = diag(6)
  ),
  list(
    ex          = "#2",
    diag        = 1.0, r = 1.0,
    description = "All entries 1. Six perfect copies of one factor; the unit-weighted composite has Spearman-Brown reliability (6&middot;1)/(1+5&middot;1) = 1 for measuring that single factor, so nFK = 1.",
    m           = matrix(1, 6, 6)
  ),
  list(
    ex          = "#3",
    diag        = 0.7, r = 0.7,
    description = "All entries 0.7. One factor measured by 6 items with diag = off-diag = 0.7. Spearman-Brown reliability of the unit composite = (6&middot;0.7)/(1+5&middot;0.7) = 0.933 &mdash; exactly matching nFK.",
    m           = matrix(0.7, 6, 6)
  ),
  list(
    ex          = "#4",
    diag        = 0.7, r = 0.0,
    description = "Six orthogonal items, each measured with reliability 0.7. Each item is its own 1-item factor; nFK = 6 &times; 0.7 = 4.2 is the sum of per-item reliabilities.",
    m           = mk(0, dg = 0.7)
  ),
  list(
    ex          = "#5",
    diag        = 1.0, r = 0.2,
    description = "Six perfectly reliable items at weak inter-correlation (r = 0.2). With diag = 1, every PC is exactly recoverable from items, so nFK = p = 6 regardless of off-diagonal magnitude. Inter-correlation is not a measurement-error discount &mdash; to register it as one, encode reliability or communality &lt; 1 on the diagonal (see #7 for the same off-diagonal structure with diag = 0.9).",
    m           = mk(0.2)
  ),
  list(
    ex          = "#6",
    diag        = 1.0, r = 0.7,
    description = "Same setup as #5 but at r = 0.7. Same principle: with diag = 1 (no measurement error), all six dimensions are exactly recoverable from the items, so nFK = 6 even at strong inter-item correlation. See #8 for the diag = 0.9 counterpart.",
    m           = mk(0.7)
  ),
  list(
    ex          = "#7",
    diag        = 0.9, r = 0.2,
    description = "Same off-diagonal structure as #5 (r = 0.2) but at diag = 0.9 &mdash; items now have 10% measurement error. Unlike #5, this matrix discounts each PC for that error, and inter-correlation now reduces nFK below p. Closed form: R&sup2;<sub>PC1</sub> = (0.9 + 5&middot;0.2)/(1 + 5&middot;0.2) = 0.95; R&sup2;<sub>k&gt;1</sub> = (0.9 &minus; 0.2)/(1 &minus; 0.2) = 0.875 each; sum = 5.325.",
    m           = mk(0.2, dg = 0.9)
  ),
  list(
    ex          = "#8",
    diag        = 0.9, r = 0.7,
    description = "Same off-diagonal structure as #6 (r = 0.7) but at diag = 0.9. Strong inter-correlation paired with imperfect reliability produces a more substantial discount than #7. Closed form: R&sup2;<sub>PC1</sub> = (0.9 + 5&middot;0.7)/(1 + 5&middot;0.7) = 0.978; R&sup2;<sub>k&gt;1</sub> = (0.9 &minus; 0.7)/(1 &minus; 0.7) = 0.667 each; sum = 4.311.",
    m           = mk(0.7, dg = 0.9)
  ),
  list(
    ex          = "#9",
    diag        = NA_real_, r = NA_real_,
    description = "Two orthogonal three-item blocks, perfectly intercorrelated within (r = 1) and uncorrelated across. Per-block Spearman-Brown = (3&middot;1)/(1+2&middot;1) = 1; nFK = 2 &times; 1 = 2.",
    m           = { x <- matrix(0, 6, 6); x[1:3, 1:3] <- 1; x[4:6, 4:6] <- 1; x }
  ),
  list(
    ex          = "#10",
    diag        = NA_real_, r = NA_real_,
    description = "Same block structure as #9 but scaled by 0.7. Per-block Spearman-Brown = (3&middot;0.7)/(1+2&middot;0.7) = 0.875; nFK = 2 &times; 0.875 = 1.75.",
    m           = { x <- matrix(0, 6, 6); x[1:3, 1:3] <- 0.7; x[4:6, 4:6] <- 0.7; x }
  ),
  list(
    ex          = "#11",
    diag        = NA_real_, r = NA_real_,
    description = "Variable A is a perfect compound: A = 0.7&middot;B + 0.3&middot;C + 0.1&middot;D, with B, C, D, E, F mutually uncorrelated unit-variance variables (so &sigma;<sup>2</sup>(A) = 0.7<sup>2</sup> + 0.3<sup>2</sup> + 0.1<sup>2</sup> = 0.59). The correlation matrix is obtained from the covariance via <code>cov2cor()</code>. Rank 5 (A is determined by B, C, D); PC6 has zero loadings and contributes 0, while the other five PCs are perfectly recovered, giving nFK = 5.",
    m           = {
      cov_m <- matrix(c(.59, .7, .3, .1, 0, 0,
                        .7,  1,  0,  0, 0, 0,
                        .3,  0,  1,  0, 0, 0,
                        .1,  0,  0,  1, 0, 0,
                        0,   0,  0,  0, 1, 0,
                        0,   0,  0,  0, 0, 1),
                      nrow = 6, byrow = TRUE)
      cov2cor(cov_m)
    }
  )
)

fmt_matrix <- function(m, vn, digits = 3) {
  dimnames(m) <- list(vn, vn)
  paste0(
    "<pre>",
    paste(capture.output(print(round(m, digits))), collapse = "\n"),
    "</pre>"
  )
}

fmt_desc <- function(ex, description) {
  sprintf(
    "<div class='nfk-desc'><span class='nfk-ex'>%s</span> %s</div>",
    ex, description
  )
}

fmt_num <- function(x) {
  if (is.na(x)) return("&mdash;")
  if (abs(x - round(x)) < 1e-6) return(sprintf("%d", as.integer(round(x))))
  sprintf("%.3f", x)
}

meff_methods <- c("nyholt", "liji", "gao", "galwey", "chen", "gao_asis", "galwey_asis")

# Unfloored versions of the five poolr::meff() formulas. poolr applies
# floor(m) before returning to give a conservative integer count for
# Bonferroni correction; we skip that step here so the continuous
# gradations distinguishing matrices remain visible. The first five use a
# diag-1 copy because their formulas assume a correlation matrix; the two
# scale-invariant methods (Gao, Galwey) are also computed on the matrix
# as supplied so the reliability-on-diagonal information is retained.
meff_raw <- function(m) {
  R <- m; diag(R) <- 1
  evs       <- eigen(R, symmetric = TRUE)$values
  evs_pos   <- pmax(evs, 0)
  k         <- length(evs)
  abs_evs   <- abs(evs) + sqrt(.Machine$double.eps)

  evs_asis  <- pmax(eigen(m, symmetric = TRUE)$values, 0)

  c(
    nyholt      = 1 + (k - 1) * (1 - var(evs) / k),
    liji        = sum(ifelse(abs_evs >= 1, 1, 0) + (abs_evs - floor(abs_evs))),
    gao         = which(cumsum(sort(evs, decreasing = TRUE)) / sum(evs) > 0.995)[1],
    galwey      = sum(sqrt(evs_pos))^2 / sum(evs_pos),
    chen        = sum(1 / apply(R, 1, function(r) sum(abs(r)^7))),
    gao_asis    = which(cumsum(sort(evs_asis, decreasing = TRUE)) / sum(evs_asis) > 0.995)[1],
    galwey_asis = sum(sqrt(evs_asis))^2 / sum(evs_asis)
  )
}

ref_results <- lapply(ref, function(r) {
  res <- tryCatch(suppressWarnings(nFK(r$m)), error = function(e) NULL)
  nfk_val <- if (is.null(res)) NA_real_ else res$nFK
  meff_vec <- tryCatch(meff_raw(r$m),
                       error = function(e) setNames(rep(NA_real_, length(meff_methods)), meff_methods))
  list(
    ex          = r$ex,
    diag        = r$diag,
    r           = r$r,
    nFK         = nfk_val,
    meff        = meff_vec,
    matrix_html = fmt_matrix(r$m, varnames),
    desc_html   = fmt_desc(r$ex, r$description)
  )
})

# Annotation table for the sweep plot below.
ref_points <- data.frame(
  ex   = vapply(ref_results, `[[`, character(1), "ex"),
  diag = vapply(ref_results, `[[`, numeric(1),   "diag"),
  r    = vapply(ref_results, `[[`, numeric(1),   "r"),
  nFK  = vapply(ref_results, `[[`, numeric(1),   "nFK"),
  stringsAsFactors = FALSE
)

get_meff <- function(name) {
  vapply(ref_results, function(r) fmt_num(r$meff[[name]]), character(1))
}

fmt_nfk <- function(x) if (is.na(x)) "&mdash;" else sprintf("%.3f", x)

df <- data.frame(
  Matrix                = vapply(ref_results, `[[`, character(1), "matrix_html"),
  `Matrix description`  = vapply(ref_results, `[[`, character(1), "desc_html"),
  nFK                   = vapply(ref_results, function(r) fmt_nfk(r$nFK), character(1)),
  nyholt                = get_meff("nyholt"),
  liji                  = get_meff("liji"),
  gao                   = get_meff("gao"),
  galwey                = get_meff("galwey"),
  chen                  = get_meff("chen"),
  `gao*`                = get_meff("gao_asis"),
  `galwey*`             = get_meff("galwey_asis"),
  check.names           = FALSE,
  stringsAsFactors      = FALSE
)

kbl <- knitr::kable(
  df, format = "html", escape = FALSE,
  align = c("l", "l", "r", "r", "r", "r", "r", "r", "r", "r"),
  table.attr = "class='nfk-ref'"
)
htmltools::HTML(as.character(kbl))
```

## Per-component reliabilities

`nFK()` returns more than the headline total. The full output includes `r2` &mdash; the per-component &omega;<sub>max</sub> vector that sums to `nFK` &mdash; alongside the `PCAloadings`. The vector is independently informative: it shows which dimensions are reliably measured and which are not. Case #11 above (variable A constructed as a perfect linear combination of B, C, D, so the matrix has rank 5) makes this explicit:

```{r r2_demo}
res11 <- nFK(ref[[11]]$m)
round(res11$r2, 3)
```

The first five components hit `r2 = 1` (perfectly recoverable from the items), and PC6 contributes exactly 0 &mdash; the rank-5 matrix cannot support a sixth orthogonal factor. The total `nFK = 5` is the count of unit r&sup2; entries; rank deficiency surfaces directly in the per-component vector.

## Connection to coefficient alpha and Spearman-Brown reliability

Each per-PC `R²_k` returned by `nFK()` admits a clean psychometric reading:

> `R²_k` is the **maximum reliability** of the optimal linear composite of items measuring the *k*-th principal component &mdash; equivalent to &omega;<sub>max</sub> (Bentler 1968; Lord 1958), the upper bound on reliability achievable by reweighting items.

When items measuring a factor are **tau-equivalent** (uniform loadings, uniform within-cluster correlation), &omega;<sub>max</sub> collapses to Cronbach's &alpha;, and for symmetric scales further to Spearman-Brown. Several reference cases above are exactly tau-equivalent within their cluster, and `nFK` lands precisely on the canonical reliability formula:

```{r verify_alpha}
verify <- data.frame(
  case      = c("#3", "#4", "#9", "#10"),
  structure = c("1 factor, 6 items, all r = 0.7, diag = 0.7",
                "6 orthogonal 1-item factors, each at reliability 0.7",
                "2 blocks of 3 items, within-block r = 1, diag = 1",
                "2 blocks of 3 items, within-block r = 0.7, diag = 0.7"),
  expected  = c("Spearman-Brown(6, 0.7) = (6&middot;0.7)/(1+5&middot;0.7) = 0.933",
                "6 &times; 0.7 = 4.200",
                "2 &times; Spearman-Brown(3, 1) = 2 &times; 1 = 2.000",
                "2 &times; Spearman-Brown(3, 0.7) = 2 &times; 0.875 = 1.750"),
  computed  = c(
    sprintf("%.3f", nFK(matrix(0.7, 6, 6))$nFK),
    sprintf("%.3f", nFK({ x <- matrix(0, 6, 6); diag(x) <- 0.7; x })$nFK),
    sprintf("%.3f", nFK({ x <- matrix(0, 6, 6); x[1:3,1:3] <- 1; x[4:6,4:6] <- 1; x })$nFK),
    sprintf("%.3f", nFK({ x <- matrix(0, 6, 6); x[1:3,1:3] <- 0.7; x[4:6,4:6] <- 0.7; x })$nFK)
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
names(verify) <- c("Case", "Structure", "Expected (closed form)", "nFK")
knitr::kable(verify, format = "html", escape = FALSE,
             caption = "Cases where nFK exactly matches the canonical reliability formula.")
```

### Why diag = 1 cases give nFK = *p*

Cases #5 and #6 (r = 0.2 and r = 0.7, both with diag = 1) both yield `nFK = 6`, regardless of off-diagonal magnitude. Under the &omega;<sub>max</sub> reading, this is the *correct* answer: with diag = 1 every item is its own 1-item perfectly-reliable composite, so the matrix encodes "I have six perfectly-measured dimensions" &mdash; even if those dimensions correlate.

To register inter-item correlation as a *measurement-information* loss, the reduction must live on the diagonal. That is the role of nFK's diagonal input: it carries reliability (or communality), and the off-diagonals describe the structure of true-score associations among items. Setting r = 0.7 with diag = 1 (case #6) asks nFK to read "I have six perfectly-measured items that happen to correlate"; setting diag &lt; 1 (as in #3, #4, #7, #8, #10) shifts the matrix toward "each item recovers less than full true-score variance," and the formula responds accordingly. The contrast is clearest between #5 vs #7 and #6 vs #8: identical off-diagonal patterns at diag = 1 give nFK = 6, while at diag = 0.9 the same matrices give 5.325 and 4.311 respectively. Cases #5 and #6 thus serve as demonstrations of the principle &mdash; not failures of the formula.

## Rotation: total invariance, per-component redistribution

`nFK()` accepts `rotate = "none"` (default) or `rotate = "varimax"`, applied to the principal-component loadings before `r2` is computed. Total `nFK` is invariant under orthogonal rotation &mdash; the same dimensional content is being re-expressed, not changed &mdash; but the per-component `r2` values redistribute. When the question is "what is the reliability of each rotated factor?" rather than "what is the dimensionality of the set?", the rotated `r2` vector is the primary output.

Two clean clusters of three items each (within-cluster *r* = 0.7, between-cluster *r* = 0, diagonal = 0.9):

```{r rotate_demo}
m_two <- matrix(0, 6, 6)
m_two[1:3, 1:3] <- 0.7
m_two[4:6, 4:6] <- 0.7
diag(m_two) <- 0.9
dimnames(m_two) <- list(varnames, varnames)

res_none <- nFK(m_two, rotate = "none")
res_vmax <- nFK(m_two, rotate = "varimax")

cat(sprintf("Total nFK (unrotated):  %.4f\n", res_none$nFK))
cat(sprintf("Total nFK (varimax):    %.4f\n\n", res_vmax$nFK))
cat("Per-component r2 (unrotated):\n"); print(round(res_none$r2, 3))
cat("\nPer-component r2 (varimax):\n");   print(round(res_vmax$r2, 3))
```

The unrotated decomposition follows the closed forms from cases #7&ndash;#8 applied within each cluster: each cluster's leading composite carries r&sup2; = (d + (k&minus;1)r)/(1 + (k&minus;1)r) = 2.3/2.4 = 0.958, and each within-cluster contrast carries (d&minus;r)/(1&minus;r) = 0.667. Varimax (with Kaiser normalization) shifts a small amount of loading mass between cluster and noise components, so the cluster components drop slightly and two of the contrast components rise &mdash; but the totals match exactly.

## Uniform-correlation sweeps

For a *p* &times; *p* matrix with constant diagonal *d* and constant off-diagonal *r*, the curves below trace nFK as *r* varies. One curve per reliability level *d* &isin; {0.1, 0.2, &hellip;, 1.0}; for each curve, *r* runs from 0 to *d* (beyond *r* = *d* the matrix is no longer positive semi-definite). A dotted curve at *d* = 0.99 (off the 0.1 grid) is included as a contrast point: with diag = 1.0, nFK stays at *p* for all *r*, but a 1% nudge below perfect reliability lets the statistic descend smoothly toward 1 as *r* &rarr; 0.99. Each panel uses its own y-axis so the curve shape is visible at every scale. Reference cases from the table above are annotated on the *p* = 6 panel.

```{r sweep, fig.width=9, fig.height=6.5}
sweep_uniform <- function(p, dg, rs) {
  out <- numeric(length(rs))
  for (i in seq_along(rs)) {
    m <- matrix(rs[i], p, p); diag(m) <- dg
    res <- tryCatch(suppressWarnings(nFK(m)), error = function(e) NULL)
    out[i] <- if (is.null(res)) NA_real_ else res$nFK
  }
  out
}

diags      <- seq(0.1, 1.0, by = 0.1)
diag_extra <- 0.99
ps         <- c(2, 6, 60, 120)
r_step     <- 0.01

df_sweep <- do.call(rbind, lapply(ps, function(p) {
  do.call(rbind, lapply(diags, function(d) {
    rs <- seq(0, d, by = r_step)
    data.frame(p = p, diag = d, r = rs, nFK = sweep_uniform(p, d, rs))
  }))
}))
df_sweep$p_lab <- factor(paste0("p = ", df_sweep$p),
                         levels = paste0("p = ", ps))

# diag = 0.99 sits off the 0.1 grid; rendered separately as a dotted line
# with per-facet labels so it isn't confused with the regular sweep.
df_sweep_extra <- do.call(rbind, lapply(ps, function(p) {
  rs <- seq(0, diag_extra, by = r_step)
  data.frame(p = p, diag = diag_extra, r = rs, nFK = sweep_uniform(p, diag_extra, rs))
}))
df_sweep_extra$p_lab <- factor(paste0("p = ", df_sweep_extra$p),
                               levels = paste0("p = ", ps))

# Label point: a position on the descending portion of each curve where
# the dotted line is clearly separated from diag = 1.0.
extra_labels <- do.call(rbind, lapply(ps, function(p) {
  sub <- df_sweep_extra[df_sweep_extra$p == p, ]
  sub[which.min(abs(sub$r - 0.95)), ]
}))
extra_labels$label <- "diag = 0.99"

ann <- transform(subset(ref_points, !is.na(r) & !is.na(diag)),
                 p     = 6L,
                 p_lab = factor("p = 6", levels = paste0("p = ", ps)))

ggplot(df_sweep, aes(x = r, y = nFK, color = diag, group = diag)) +
  geom_line(alpha = 0.85, linewidth = 0.6) +
  geom_line(data = df_sweep_extra,
            aes(x = r, y = nFK, group = diag),
            inherit.aes = FALSE,
            color = "grey20", linetype = "dotted", linewidth = 0.7) +
  geom_label(data = extra_labels,
             aes(x = r, y = nFK, label = label),
             inherit.aes = FALSE,
             hjust = 1.0, vjust = 1.4, size = 3,
             label.size = 0, label.padding = unit(0.1, "lines"),
             fill = alpha("white", 0.7), show.legend = FALSE) +
  geom_point(data = ann, size = 2.6, shape = 21, fill = "white", stroke = 1) +
  geom_label(data = ann, aes(label = ex),
             vjust = -0.4, hjust = -0.1, size = 3.3,
             label.size = 0, label.padding = unit(0.1, "lines"),
             fill = alpha("white", 0.7), show.legend = FALSE) +
  facet_wrap(~ p_lab, scales = "free_y") +
  scale_color_viridis_c(
    name      = "diag\n(reliability)",
    limits    = c(0.1, 1.0),
    breaks    = diags,
    direction = -1,
    option    = "viridis"
  ) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.2),
                     expand = expansion(mult = c(0.02, 0.08))) +
  scale_y_continuous(expand = expansion(mult = c(0.04, 0.10))) +
  expand_limits(y = 0) +
  labs(
    x     = "Off-diagonal correlation (r)",
    y     = "nFK",
    title = "nFK across uniform correlation matrices, by matrix size and reliability"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        strip.text       = element_text(face = "bold"))
```
