---
title: "How much of a trait's stability runs through experience?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How much of a trait's stability runs through experience?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
```

```{r setup}
library(fancyr)
```

## The question

People who are high on a trait at one time tend to be high on it later. That
**stability** can come about in several ways. One of them runs through
experience: a trait can lead people into certain experiences (*selection*),
and those experiences can then reinforce the same trait (*socialization*).
Roberts, Caspi and Moffitt (2003) called this the *corresponsive principle*.
When it holds, part of a trait's continuity is carried by the experiences
people select into.

`stabilityPaths()` estimates that part directly. For each item measured at two
waves, it splits the total stability coefficient -- the prediction of Time 2
scores from Time 1 scores -- into additive pieces:

| pathway | route | meaning |
|---|---|---|
| **mediated** | `Y1 -> X -> Y2` | stability carried by an experience X: selection × socialization |
| **confounded** | `Y1 <-> C -> Y2` | stability due to a stable third variable C |
| **residual** | `Y1 -> Y2` | stability not accounted for by either |

The pieces add up to the total exactly. Dividing each by the total gives its
**share** of the item's stability.

## The data

`stabilitySim` is a simulated study where the right answer is known. Six
personality items were measured a year apart, and about 30% of people took a
leadership role in between. Leadership was chosen by dominant (and, more
weakly, sociable) people and increased those traits in turn. It also reduced
anxiety, but anxiety played no part in who led. Socioeconomic status (`ses`)
is related to several traits and to leadership. See `?stabilitySim` for
details.

The raw pieces are separate files, as they usually are in practice: a Time 1
file, a Time 2 file, and a record listing only the people who led.

```{r}
str(stabilitySim$T1)
head(stabilitySim$experience, 3)
```

`stabilityData()` puts them together. Items measured at both waves get `[T1]`
and `[T2]` suffixes. Columns found in only one wave are dropped by default;
`keep = "T1"` keeps the Time 1 ones, such as `ses`, under their own names. Extra
files are joined on the ID, and `fill` codes everyone missing from the
leadership file as 0.

```{r}
d <- stabilityData(stabilitySim$T1, stabilitySim$T2, stabilitySim$experience,
                   keep = "T1", fill = list(leader = 0), date = "date")
names(d)
summary(d$interval_days)
```

### Who is in the sample?

About a fifth of participants never returned for Time 2, and they were not a
random fifth: dropout was more likely for people lower in `ses` and in
`organized`. By default `stabilityData()` **keeps** those people:

```{r}
c(everyone = nrow(d), returned = sum(!is.na(d[["organized[T2]"]])))
```

That's deliberate. `stabilityPaths()` estimates its models by *full
information maximum likelihood* (FIML), which uses every value each person
has. Someone seen only at Time 1 still tells us about the Time 1 distribution
and about how Time 1 scores relate to leadership and `ses`.

Dropping them (listwise deletion, `join = "inner"`) is unbiased only if
dropout is *missing completely at random*: unrelated to anything measured or
not. FIML only needs dropout to be *missing at random*: it may depend on
observed things like Time 1 scores, as it does here, as long as it doesn't
further depend on the unobserved Time 2 score itself (Rubin, 1976; Schafer &
Graham, 2002; Enders, 2010). The *Missing data* section of `?stabilityData`
has more detail and references.

## A first decomposition

```{r}
obs <- stabilityPaths(d, X = "leader", controls = "ses")
obs
```

Each cell is an estimate, followed by its share of the item's total
stability. In the standardized metric (the default), the total is the retest
correlation.

## Adjusting for retest reliability

These are single items, and a single answer reflects the particular occasion
as well as where the person usually stands. That matters here in a specific
way. The model asks whether leadership predicts Time 2 scores *over and above*
Time 1 scores. A single Time 1 answer controls only partly for the person's
expected score around Time 1. The rest of that expected score gets credited
to leadership, because dominant people both led and stayed dominant. That
inflates the socialization path and with it the mediated share. It also
deflates the residual stability, since occasion-specific influences at either
wave weaken the retest correlation.

If you know an item's reliability, ideally a retest correlation over a few
weeks (see *Choosing a reliability* below), supply it. Each item is then
modelled as a latent variable measured with that reliability:

```{r}
stabilitySim$reliability
lat <- stabilityPaths(d, X = "leader", controls = "ses",
                      reliability = stabilitySim$reliability)
lat
```

Latent items are drawn as ellipses in the path diagram, labelled with the
reliability assumed. The total is now the stability of the *expected scores*
(the retest correlation adjusted for the items' reliabilities), and the
components still sum to it exactly.

Because the data are simulated, we can check both analyses against the truth:

```{r}
tr  <- stabilitySim$truth
key <- function(p) paste(p$item, p$path)
cmp <- data.frame(
  item     = obs$paths$item,
  path     = obs$paths$path,
  truth    = tr$est[match(key(obs$paths), key(tr))],
  observed = round(obs$paths$est, 3),
  latent   = round(lat$paths$est, 3)
)
cmp[cmp$item %in% c("dominant", "sociable", "anxious"), ]
```

The latent analysis lands close to the true values throughout. The observed
analysis understates every item's total stability, and it overstates the share
of `dominant`'s stability that runs through leadership:

```{r}
share <- function(x, it) x$paths$share[x$paths$item == it & x$paths$type == "mediated"]
round(c(truth    = tr$share[tr$item == "dominant" & tr$type == "mediated"],
        observed = share(obs, "dominant"),
        latent   = share(lat, "dominant")), 3)
```

## Reading the results

`summary()` adds standard errors, confidence intervals and p-values. It also
shows the two structural paths whose product *is* each mediated pathway:
selection (`leader ~ Y1`) and socialization (`Y2 ~ leader`).

```{r}
s <- summary(lat)
s$structural
```

The `anxious` row shows why the decomposition is informative beyond the
socialization effect alone. Leadership clearly *changes* anxiety (a sizeable
negative `Y2 ~ leader`), yet it carries none of anxiety's stability. With no
selection on anxiety, there is nothing for the socialization effect to
multiply. The trait–experience relationship is real, but it is
*non-corresponsive* in the sense of Wood and Harms (2017): the experience
changes a trait that did not select people into it, so it doesn't contribute
to that trait's continuity.

## Plots

With no arguments, `plot()` compares items. Each bar is an item's total
stability, divided into its pathways:

```{r}
plot(lat)
```

With `item`, it draws that item's path diagram:

```{r, fig.height = 5}
plot(lat, item = "dominant")
```

## Choosing a reliability

An item given a reliability stands for the person's *expected score* on it
over some interval *t*: roughly, the average of the answers they would give if
asked repeatedly across that span. Nothing assumes the expected score is
fixed. It can change between the waves, which is exactly what the model
studies. The reliability to supply is the share of variance in a single
answer that is shared with the expected score over *t*.

A retest correlation over an interval of about *t* estimates this directly
(Wood et al., 2023). Whatever contributes to both answers, meaning whatever
persists over *t*, is part of the expected score. Influences that come and go
within *t*, such as mood, the day's circumstances, or remembering an answer
given a moment ago, still shape each answer systematically. They just aren't
part of the expected score over *t*. As *t* stretches from an hour to a day,
a week, a month or a year, more of what lies behind an answer has the chance
to change, and retest correlations fall (Watson, 2004).

So choosing *t* decides what the adjusted results describe:

- **Short *t*.** With a same-session retest, or one a few days or a week
  apart, nearly everything behind an answer persists. The reliability is high
  and the adjustment small, and the adjusted stability still describes scores
  tied to the particular occasion.
- **Long *t*.** As *t* approaches the gap between the waves, the change the
  model is meant to study gets folded into the reliability, and the
  adjustment does too much. If you use the Time 1–Time 2 retest correlation
  itself, the adjusted stability is 1 by construction.
- **In between.** Long enough that what is specific to an occasion has turned
  over, but short relative to the interval being studied. For waves a year
  apart, retest correlations over two to four weeks are a reasonable target.

Internal consistency (alpha) fits the same logic, because its items are
answered at different moments too, typically about a minute apart. Alpha is
often called an underestimate of reliability. What it underestimates is the
retest reliability of those items *over that short interval*. If two items
answered about a minute apart give an alpha of .80, each item's own retest
correlation over a minute (averaged, say, as a geometric mean) should exceed
.80. Over longer intervals such as two weeks, observed retest correlations are
regularly *lower* than alpha (McCrae et al., 2011; Chmielewski & Watson,
2009). So alpha isn't a suitable value for waves months or years apart, and a
single item has no alpha anyway.

Suitable retest values often don't exist for the items you have. Published
retest correlations for similar items or measures can suggest a range, and
the sensitivity analysis below shows whether the conclusions hold across it.

## How much does the assumed reliability matter?

Standard errors treat the supplied reliability as known, but it's an estimate
too, and conclusions can depend on it. `reliabilitySensitivity()` refits the
analysis across a grid of reliabilities. A value of 1 is the unadjusted
analysis:

```{r, fig.width = 8, fig.height = 4}
sens <- reliabilitySensitivity(obs, rel = seq(.5, 1, .1))
plot(sens)
```

The mediated share for `dominant` shrinks steadily as the assumed reliability
falls. Points drop out where an assumed reliability is too low for the data
(adjusted stability above 1), which is itself a useful
bound on what reliability is plausible.

## Interpreting shares, and some cautions

- **Shares are shares of stability**, not of variance in Time 2 scores. A
  small mediated share can still reflect a large experience effect when
  selection is weak, as the `anxious` example shows.
- **Timing.** The mediated pathway treats the experience as falling between
  the waves. If it was measured at Time 2, or spans the whole interval, that
  order is assumed rather than observed (Maxwell & Cole, 2007).
- **Unmeasured confounding.** Controls address only the confounders you
  measure. A stable variable left out of `controls` inflates both the
  socialization path and the residual path.
- **Totals near zero** make shares unstable. Read shares alongside the
  estimates.
- **Binary experiences** such as `leader` should not be given a reliability:
  a record of whether something happened isn't a fallible measure of a
  continuous trait.

## Real data: traits and social power

*This section will use data from Wood and Harms (2017) once it's added to
the package.*

## References

Chmielewski, M., & Watson, D. (2009). What is being assessed and why it
matters: The impact of transient error on trait research. *Journal of
Personality and Social Psychology, 97*(1), 186--202.

Enders, C. K. (2010). *Applied missing data analysis*. Guilford Press.

Maxwell, S. E., & Cole, D. A. (2007). Bias in cross-sectional analyses of
longitudinal mediation. *Psychological Methods, 12*(1), 23--44.

McCrae, R. R., Kurtz, J. E., Yamagata, S., & Terracciano, A. (2011).
Internal consistency, retest reliability, and their implications for
personality scale validity. *Personality and Social Psychology Review,
15*(1), 28--50.

Roberts, B. W., Caspi, A., & Moffitt, T. E. (2003). Work experiences and
personality development in young adulthood. *Journal of Personality and Social
Psychology, 84*(3), 582--593.

Rubin, D. B. (1976). Inference and missing data. *Biometrika, 63*(3), 581--592.

Schafer, J. L., & Graham, J. W. (2002). Missing data: Our view of the state of
the art. *Psychological Methods, 7*(2), 147--177.

Watson, D. (2004). Stability versus change, dependability versus error:
Issues in the assessment of personality over time. *Journal of Research in
Personality, 38*(4), 319--350.

Wood, D., & Harms, P. D. (2017). Evidence of non-corresponsive causal
relationships between personality traits and social power over time.
*Personality and Social Psychology Bulletin, 43*(1), 33--45.

Wood, D., Lowman, G. H., Armstrong, B. F., III, & Harms, P. D. (2023). Using
retest-adjusted correlations as indicators of the semantic similarity of
items. *Journal of Personality and Social Psychology, 125*(2), 437--454.
https://doi.org/10.1037/pspp0000441
