---
title: "How much of a trait's stability runs through experience?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How much of a trait's stability runs through experience?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
```

```{r setup}
library(fancyr)
```

## The question

People who are high on a trait at one time tend to be high on it later. That
**stability** can come about in several ways. One of them runs through
experience: a trait can lead people into certain experiences (*selection*),
and those experiences can then reinforce the same trait (*socialization*).
Roberts, Caspi and Moffitt (2003) called this the *corresponsive principle*.
When it holds, part of a trait's continuity is carried by the experiences
people select into.

`stabilityPaths()` estimates that part directly. For each item measured at two
waves, it splits the total stability coefficient -- the prediction of Time 2
scores from Time 1 scores -- into additive pieces:

| pathway | route | meaning |
|---|---|---|
| **mediated** | `Y1 -> X -> Y2` | stability carried by an experience X: selection × socialization |
| **confounded** | `Y1 <-> C -> Y2` | stability due to a stable third variable C |
| **residual** | `Y1 -> Y2` | stability not accounted for by either |

The pieces add up to the total exactly. Dividing each by the total gives its
**share** of the item's stability.

## The data

`stabilitySim` is a simulated study where the right answer is known. Six
personality items were measured a year apart, and about 30% of people took a
leadership role in between. Leadership was chosen by dominant (and, more
weakly, sociable) people and increased those traits in turn. It also reduced
anxiety, but anxiety played no part in who led. Socioeconomic status (`ses`)
is related to several traits and to leadership. See `?stabilitySim` for
details.

The raw pieces are separate files, as they usually are in practice: a Time 1
file, a Time 2 file, and a record listing only the people who led.

```{r}
str(stabilitySim$T1)
head(stabilitySim$experience, 3)
```

`stabilityData()` puts them together. Items measured at both waves get `[T1]`
and `[T2]` suffixes. Other Time 1 columns such as `ses` keep their names. Extra
files are joined on the ID, and `fill` codes everyone missing from the
leadership file as 0.

```{r}
d <- stabilityData(stabilitySim$T1, stabilitySim$T2, stabilitySim$experience,
                   fill = list(leader = 0), date = "date")
names(d)
summary(d$interval_days)
```

### Who is in the sample?

About a fifth of participants never returned for Time 2, and they were not a
random fifth: dropout was more likely for people lower in `ses` and in
`organized`. By default `stabilityData()` **keeps** those people:

```{r}
c(everyone = nrow(d), returned = sum(!is.na(d[["organized[T2]"]])))
```

That's deliberate. `stabilityPaths()` estimates its models by *full
information maximum likelihood* (FIML), which uses every value each person
has. Someone seen only at Time 1 still tells us about the Time 1 distribution
and about how Time 1 scores relate to leadership and `ses`.

Dropping them (listwise deletion, `join = "inner"`) is unbiased only if
dropout is *missing completely at random*: unrelated to anything measured or
not. FIML only needs dropout to be *missing at random*: it may depend on
observed things like Time 1 scores, as it does here, as long as it doesn't
further depend on the unobserved Time 2 score itself (Rubin, 1976; Schafer &
Graham, 2002; Enders, 2010). The *Missing data* section of `?stabilityData`
has more detail and references.

## A first decomposition

```{r}
obs <- stabilityPaths(d, X = "leader", controls = "ses")
obs
```

Each cell is an estimate, followed by its share of the item's total
stability. In the standardized metric (the default), the total is the retest
correlation.

## Correcting for measurement error

These are single items, and single items are noisy. That matters here in a
specific way. The model asks whether leadership predicts Time 2 scores *over
and above* Time 1 scores. If Time 1 scores are unreliable, "adjusting for Time
1" only partly adjusts for the true trait. The unadjusted remainder of the
trait gets credited to leadership, because dominant people both led and stayed
dominant. That inflates the socialization path and with it the mediated share.
It also deflates the residual stability, since noise at either wave weakens
the retest correlation.

If you know an item's reliability, for example from a short-interval retest
where the trait itself shouldn't have changed, supply it. Each item is then
modelled as a latent variable measured with that reliability:

```{r}
stabilitySim$reliability
lat <- stabilityPaths(d, X = "leader", controls = "ses",
                      reliability = stabilitySim$reliability)
lat
```

Latent items are drawn as ellipses in the path diagram, labelled with the
reliability assumed. The total is now the retest correlation *corrected for
unreliability*, and the components still sum to it exactly.

Because the data are simulated, we can check both analyses against the truth:

```{r}
tr  <- stabilitySim$truth
key <- function(p) paste(p$item, p$path)
cmp <- data.frame(
  item     = obs$paths$item,
  path     = obs$paths$path,
  truth    = tr$est[match(key(obs$paths), key(tr))],
  observed = round(obs$paths$est, 3),
  latent   = round(lat$paths$est, 3)
)
cmp[cmp$item %in% c("dominant", "sociable", "anxious"), ]
```

The latent analysis lands close to the true values throughout. The observed
analysis understates every item's total stability, and it overstates the share
of `dominant`'s stability that runs through leadership:

```{r}
share <- function(x, it) x$paths$share[x$paths$item == it & x$paths$type == "mediated"]
round(c(truth    = tr$share[tr$item == "dominant" & tr$type == "mediated"],
        observed = share(obs, "dominant"),
        latent   = share(lat, "dominant")), 3)
```

## Reading the results

`summary()` adds standard errors, confidence intervals and p-values. It also
shows the two structural paths whose product *is* each mediated pathway:
selection (`leader ~ Y1`) and socialization (`Y2 ~ leader`).

```{r}
s <- summary(lat)
s$structural
```

The `anxious` row shows why the decomposition is informative beyond the
socialization effect alone. Leadership clearly *changes* anxiety (a sizeable
negative `Y2 ~ leader`), yet it carries none of anxiety's stability. With no
selection on anxiety, there is nothing for the socialization effect to
multiply. The trait–experience relationship is real, but it is
*non-corresponsive* in the sense of Wood and Harms (2017): the experience
changes a trait that did not select people into it, so it doesn't contribute
to that trait's continuity.

## Plots

With no arguments, `plot()` compares items. Each bar is an item's total
stability, divided into its pathways:

```{r}
plot(lat)
```

With `item`, it draws that item's path diagram:

```{r, fig.height = 5}
plot(lat, item = "dominant")
```

## How much does the assumed reliability matter?

Standard errors treat the supplied reliability as known, but it's an estimate
too, and conclusions can depend on it. `reliabilitySensitivity()` refits the
analysis across a grid of reliabilities. A value of 1 is the uncorrected
analysis:

```{r, fig.width = 8, fig.height = 4}
sens <- reliabilitySensitivity(obs, rel = seq(.5, 1, .1))
plot(sens)
```

The mediated share for `dominant` shrinks steadily as more of each item's
variance is treated as error. Points drop out where an assumed reliability is
too low for the data (corrected stability above 1), which is itself a useful
bound on what reliability is plausible.

## Interpreting shares, and some cautions

- **Shares are shares of stability**, not of variance in Time 2 scores. A
  small mediated share can still reflect a large experience effect when
  selection is weak, as the `anxious` example shows.
- **Timing.** The mediated pathway treats the experience as falling between
  the waves. If it was measured at Time 2, or spans the whole interval, that
  order is assumed rather than observed (Maxwell & Cole, 2007).
- **Unmeasured confounding.** Controls address only the confounders you
  measure. A stable variable left out of `controls` inflates both the
  socialization path and the residual path.
- **Totals near zero** make shares unstable. Read shares alongside the
  estimates.
- **Binary experiences** such as `leader` should not be given a reliability:
  a record of whether something happened isn't a fallible measure of a
  continuous trait.

## Real data: traits and social power

*This section will use data from Wood and Harms (2017) once it's added to
the package.*

## References

Enders, C. K. (2010). *Applied missing data analysis*. Guilford Press.

Maxwell, S. E., & Cole, D. A. (2007). Bias in cross-sectional analyses of
longitudinal mediation. *Psychological Methods, 12*(1), 23--44.

Roberts, B. W., Caspi, A., & Moffitt, T. E. (2003). Work experiences and
personality development in young adulthood. *Journal of Personality and Social
Psychology, 84*(3), 582--593.

Rubin, D. B. (1976). Inference and missing data. *Biometrika, 63*(3), 581--592.

Schafer, J. L., & Graham, J. W. (2002). Missing data: Our view of the state of
the art. *Psychological Methods, 7*(2), 147--177.

Wood, D., & Harms, P. D. (2017). Evidence of non-corresponsive causal
relationships between personality traits and social power over time.
*Personality and Social Psychology Bulletin, 43*(1), 33--45.
