---
title: "Stability decomposition: traits and social power"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stability decomposition: traits and social power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, out.width = "100%",
                      dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png",
                      dpi = 96, fig.retina = 2)
```

```{r setup}
library(fancyr)
```

## The question

People who are high on a trait at one time tend to be high on it later. That
**stability** can come about in several ways. One of them runs through
experience: a trait can lead people into certain experiences (*selection*),
and those experiences can then reinforce the same trait (*socialization*).
Roberts, Caspi and Moffitt (2003) called this the *corresponsive principle*.
When it holds, part of a trait's continuity is carried by the experiences
people select into.

`stabilityPaths()` estimates that part directly. For each item measured at two
waves, it splits the total stability coefficient -- the prediction of Time 2
scores from Time 1 scores -- into additive pieces:

| pathway | route | meaning |
|---|---|---|
| **mediated** | `Y1 -> X -> Y2` | stability carried by an experience X: selection × socialization |
| **confounded** | `Y1 <-> C -> Y2` | stability due to a third variable C |
| **residual** | `Y1 -> Y2` | stability not accounted for by either |

The pieces add up to the total exactly. Dividing each by the total gives its
**share** of the item's stability.

## The study

This vignette works through the study behind Wood and Harms (2017). Members of
seven fraternities and sororities were surveyed twice, about a year apart. At
each wave they described themselves on 59 trait adjectives, twice over: once
as they see themselves **in general** (their *general identity*), and once as
they see themselves **as a member of their organization** (their *role
identity*). They also rated how much influence every member they knew had
within the organization. A person's **social power** is the average of the
influence ratings they received, typically from 20 to 50 other members, so it
is a very dependable estimate.

Wood and Harms asked two questions about traits and power over time:

| path | question | model |
|---|---|---|
| **A** | Do traits predict *changes* in power? | Time 2 power on Time 1 trait, controlling for Time 1 power |
| **B** | Does power predict *changes* in traits? | Time 2 trait on Time 1 power, controlling for Time 1 trait |

The answer was that the two directions differ: the relationship is
*non-corresponsive*. Communal traits predicted gains in power (Path A), while
power predicted gains in agentic role identities, such as seeing oneself as
powerful and not shy (Path B).

Both questions are stability decompositions:

- **Path B.** Treat power as the "experience". Members who see themselves as,
  say, powerful tend to *have* power (selection), and power predicts how they
  see themselves a year later (socialization, which is Path B). The product
  is the part of the trait's stability that runs through power: the mediated
  pathway, with `X = "power[T1]"`.
- **Path A.** The roles swap. Power is the item, and a trait measured at
  Time 1 is a control. The trait's link to Time 2 power, controlling for
  Time 1 power, is Path A.

## The data

`powerTraits` is a list. `T1` and `T2` hold each wave's data, one row per
person, including the survey `date`. `people` holds fixed information about
each person. `items` gives the agency and communion content of each
adjective, as rated by judges.

```{r}
str(powerTraits, max.level = 1)
powerTraits$T1[1:3, 1:8]
head(powerTraits$people, 3)
```

Each wave has `power`, the 59 general-identity adjectives (`Kind`,
`Dominant`, ...), the same adjectives as role identities (`Kind_role`, ...),
and four profile scores: `communion`, `agency`, `communion_role` and
`agency_role`. A profile score is the within-person correlation between a
person's ratings on the 59 adjectives and the adjectives' communion (or
agency) content. It is high for someone who endorses communal adjectives and
rejects uncommunal ones. See `?powerTraits` for details.

`stabilityData()` puts the pieces together. Items measured at both waves,
named in `commonItems`, get `[T1]` and `[T2]` suffixes. Other columns found in
only one wave are dropped by default, with a message; `keep = "T1"` or
`keep = "all"` retain them. Further files, such as `people`, are joined on the
ID. Given the name of a date column, `stabilityData()` also computes each
person's `interval_days`.

```{r}
pt <- powerTraits
adj <- pt$items$item
d <- stabilityData(pt$T1, pt$T2, pt$people, date = "date",
                   commonItems = c("power", "communion", "agency",
                                   "communion_role", "agency_role",
                                   adj, paste0(adj, "_role")))
summary(d$interval_days)
```

Organizations were surveyed on different schedules, so the interval between
waves ranges from about 11 to 18 months.

### Who is in the sample?

About half the Time 1 participants never answered the Time 2 survey, mostly
because seniors graduated. By default, `stabilityData()` **keeps** them:

```{r}
c(everyone = nrow(d),
  selfRatingsBothWaves = sum(!is.na(d[["communion[T1]"]]) &
                             !is.na(d[["communion[T2]"]])))
```

That's deliberate. `stabilityPaths()` estimates its models by *full
information maximum likelihood* (FIML), which uses every value each person
has. Someone seen only at Time 1 still tells us about the Time 1 distribution
and about how Time 1 scores relate to power and the controls. Here that
matters doubly, because power comes from other members' ratings: many people
who skipped the Time 2 survey still have Time 2 power.

Dropping incomplete cases (listwise deletion, `join = "inner"`) is unbiased
only if dropout is *missing completely at random*: unrelated to anything,
measured or not. FIML only needs dropout to be *missing at random*: it may
depend on observed things like Time 1 scores, as long as it doesn't further
depend on the unobserved Time 2 score itself (Rubin, 1976; Schafer & Graham,
2002; Enders, 2010). The *Missing data* section of `?stabilityData` has more
detail and references.

### Controls: organization and tenure

Organizations differ in their average power and in how their members see
themselves, and Wood and Harms controlled for organization in every analysis
with dummy codes. We do the same: six indicator variables, with organization A
as the reference, join the controls, along with `tenure`, the years a person
had been a member at Time 1.

```{r}
orgs <- paste0("org", LETTERS[2:7])
d[orgs] <- lapply(LETTERS[2:7], function(o) as.numeric(d$org == o))
ctrl <- c("tenure", orgs)
```

Organizations are single-sex, so the dummies also control for gender. Each
dummy is an ordinary control with its own confounded pathway. To keep the
printed tables readable, `print()` and `plot()` can combine them with
`pool = list(organization = orgs)`. This affects the display only: the model,
its degrees of freedom and `summary()` are unchanged, and because pathways
add, the combined column still sums with the others to the total.

## A first decomposition: power and role identities (Path B)

We start with role identities that Wood and Harms found to be related to
power, plus `Relaxed_role`, which will make a point below:

```{r}
roleItems <- c("Powerful_role", "Persuasive_role", "Assertive_role",
               "Shy_role", "Relaxed_role")
obs <- stabilityPaths(d, items = roleItems, X = "power[T1]", controls = ctrl)
print(obs, pool = list(organization = orgs))
```

Each cell is an estimate, followed by its share of the item's total
stability. In the standardized metric (the default), the total is the retest
correlation, here over about a year.

For `Powerful_role`, about a quarter of the item's stability runs through
power. Members who saw themselves as powerful in their organization had more
power at Time 1, and having power predicted seeing themselves as still more
powerful a year later. `Shy_role` shows the same pattern in reverse.

Power is measured at Time 1, at the same time as the Time 1 self-ratings,
rather than between the waves. The mediated pathway therefore treats a
concurrent association as selection: the direction of that first link is
assumed, not observed (see the cautions at the end).

## Reading the results

`summary()` adds standard errors, confidence intervals and p-values. It also
shows the two structural paths whose product *is* each mediated pathway:
selection (`power[T1] ~ Y1`) and socialization (`Y2 ~ power[T1]`, which is
Path B).

```{r}
s <- summary(obs)
s$structural
```

The `Relaxed_role` row shows why the decomposition says more than the
socialization effect alone. Power clearly *changes* how relaxed people feel in
their role: a sizeable negative `Y2 ~ power[T1]`. Yet power carries none of
that item's stability, because how relaxed people were played no part in who
had power. With no selection, there is nothing for the socialization effect
to multiply. This is non-corresponsiveness in the sense of Wood and Harms:
the experience changes a trait that didn't select people into it, so it adds
nothing to that trait's continuity.

Does power affect general identities the same way? Role experiences should
affect role identities first (Wood & Roberts, 2006):

```{r}
genItems <- sub("_role$", "", roleItems)
gen <- stabilityPaths(d, items = genItems, X = "power[T1]", controls = ctrl)
data.frame(adjective = genItems,
           role      = round(s$structural[["Y2 ~ power[T1]"]], 3),
           general   = round(summary(gen)$structural[["Y2 ~ power[T1]"]], 3))
```

Except for `Persuasive`, the effects of power are much weaker on how people
see themselves in general.

## Plots

With no arguments, `plot()` compares items. Each bar is an item's total
stability, divided into its pathways:

```{r}
plot(obs, pool = list(organization = orgs))
```

With `item`, it draws that item's path diagram. `show_controls = FALSE` leaves
the controls out of the drawing; the estimates are still adjusted for them.

```{r, fig.height = 5}
plot(obs, item = "Powerful_role", show_controls = FALSE)
```

## Adjusting for retest reliability

These are single items, and a single answer reflects the particular occasion
as well as where the person usually stands. That matters here in a specific
way. The model asks whether power predicts Time 2 self-ratings *over and
above* Time 1 self-ratings. A single Time 1 answer only partly controls for
the person's expected score around Time 1. The rest of that expected score
can be credited to power, whenever power is related to it, which inflates the
socialization path and with it the mediated share. Occasion-specific
influences at either wave also weaken the retest correlation. Wood and Harms
raised exactly this concern about their Path B estimates.

If you know an item's reliability, ideally a retest correlation over a few
weeks (see *Choosing a reliability* below), supply it. Each item is then
modelled as a latent variable measured with that reliability. We have no
short-interval retest correlations for these adjectives, so we use .75 as a
middle value and check a range below. Power gets its own value: because it
averages many raters, the reliability of each person's average is about .93
(the mean across organizations, estimated from how much raters agreed).

```{r}
rel <- c(setNames(rep(.75, length(roleItems)), roleItems), "power[T1]" = .93)
lat <- stabilityPaths(d, items = roleItems, X = "power[T1]", controls = ctrl,
                      reliability = rel)
print(lat, pool = list(organization = orgs))
```

Latent items are drawn as ellipses in the path diagram, labelled with the
reliability modeled. The total is now the stability of the *expected scores*
(the retest correlation adjusted for the items' reliabilities), and the
components still sum to it exactly.

```{r}
data.frame(item     = roleItems,
           observed = round(s$structural[["Y2 ~ power[T1]"]], 3),
           latent   = round(summary(lat)$structural[["Y2 ~ power[T1]"]], 3))
```

Here, the concern doesn't materialize: the socialization paths change little.
Two adjustments pull in opposite directions. Modelling the Time 1 self-rating
as a latent variable controls more completely for where people usually
stand, which shrinks the path. Modelling power as a latent variable
strengthens its estimated effect. For these items the two roughly cancel. The
totals rise, as they should once occasion-specific influences are set aside.

## Choosing a reliability

An item given a reliability stands for the person's *expected score* on it
over some interval *t*: roughly, the average of the answers they would give if
asked repeatedly across that span. Nothing assumes the expected score is
fixed. It can change between the waves, which is exactly what the model
studies. The reliability to supply is the share of variance in a single
answer that is shared with the expected score over *t*.

A retest correlation over an interval of about *t* estimates this directly
(Wood et al., 2023). Whatever contributes to both answers, meaning whatever
persists over *t*, is part of the expected score. Influences that come and go
within *t*, such as mood, the day's circumstances, or remembering an answer
given a moment ago, still shape each answer systematically. They just aren't
part of the expected score over *t*. As *t* stretches from an hour to a day,
a week, a month or a year, more of what lies behind an answer has the chance
to change, and retest correlations fall (Watson, 2004).

So choosing *t* decides what the adjusted results describe:

- **Short *t*.** With a same-session retest, or one a few days or a week
  apart, nearly everything behind an answer persists. The reliability is high
  and the adjustment small, and the adjusted stability still describes scores
  tied to the particular occasion.
- **Long *t*.** As *t* approaches the gap between the waves, the change the
  model is meant to study gets folded into the reliability, and the
  adjustment does too much. If you use the Time 1–Time 2 retest correlation
  itself, the adjusted stability is 1 by construction.
- **In between.** Long enough that what is specific to an occasion has turned
  over, but short relative to the interval being studied. For waves a year
  apart, retest correlations over two to four weeks are a reasonable target.

Internal consistency (alpha) fits the same logic, because its items are
answered at different moments too, typically about a minute apart. Alpha is
often called an underestimate of reliability. What it underestimates is the
retest reliability of those items *over that short interval*. If two items
answered about a minute apart give an alpha of .80, each item's own retest
correlation over a minute (averaged, say, as a geometric mean) should exceed
.80. Over longer intervals such as two weeks, observed retest correlations are
regularly *lower* than alpha (McCrae et al., 2011; Chmielewski & Watson,
2009). So alpha isn't a suitable value for waves months or years apart, and a
single item has no alpha anyway.

The reliability of power is a different kind of number. It describes how
much a person's average would change with a different set of raters, not
with a different occasion. Over a few weeks, few members' power is likely to
shift much, so it serves as a reasonable approximation here, but the two
quantities are not the same.

Suitable retest values often don't exist for the items you have. Published
retest correlations for similar items or measures can suggest a range, and
the sensitivity analysis below shows whether the conclusions hold across it.

## How much does the modeled reliability matter?

Standard errors treat the supplied reliability as known, but it's an estimate
too, and conclusions can depend on it. Often the question is not how
stability divides into shares, but whether the effects behind it -- the
selection effect (`X ~ Y1`) and the change effect (`Y2 ~ X`) -- are sizeable
and significant, and whether they stay that way as the items are modeled as
less reliable. `reliabilitySensitivity()` refits the analysis across a grid
of reliabilities for the items, keeping power's value fixed, and records
these effects with their standard errors at each value. A value of 1 is the
analysis without adjusting the items.

We apply it to the measures Wood and Harms emphasized: the communion and
agency profiles of people's general and role identities. The unadjusted
change effects are this analysis's version of Path B in their Table 1.

```{r}
profiles <- c("communion", "agency", "communion_role", "agency_role")
prof <- stabilityPaths(d, items = profiles, X = "power[T1]", controls = ctrl,
                       reliability = c("power[T1]" = .93))
sens <- reliabilitySensitivity(prof, rel = seq(.7, 1, .1))
sens
```

```{r, fig.width = 8, fig.height = 4.5}
plot(sens)
```

Filled points have p < .05, and bars are 95% confidence intervals. Without
adjustment (reliability 1), power predicts no significant change in any
profile. The largest change effect is for the agency of role identities,
about .11, the same value Wood and Harms reported for this path. Selection
effects, the concurrent link between a profile and power, are significant
throughout and grow as the profiles are modeled as less reliable. The change
effects instead shrink toward zero. So the conclusion that power does not
change people's overall agency or communion holds up, and becomes firmer,
under adjustment.

The profile scores are quite stable over the year (retest correlations near
.6), and modeled reliabilities of .5 or .6 give inadmissible solutions
(adjusted stability above 1). That is why the grid starts at .7: an
inadmissible value is itself a useful bound on what reliability is plausible.

## Selection and change effects across all adjectives

Profiles summarize many adjectives at once. To see the adjectives
individually, `plot(type = "effects")` places each item by its selection
effect (horizontal) and change effect (vertical). Here are all 59 role
identities, first without adjustment, then modeling each adjective's
reliability as .70:

```{r}
roleAll <- paste0(adj, "_role")
effObs <- stabilityPaths(d, items = roleAll, X = "power[T1]", controls = ctrl)
effLat <- stabilityPaths(d, items = roleAll, X = "power[T1]", controls = ctrl,
                         reliability = c(setNames(rep(.7, length(roleAll)), roleAll),
                                         "power[T1]" = .93))
eff <- function(x, col) summary(x)$structural[[col]]
xl <- range(0, eff(effObs, "power[T1] ~ Y1"), eff(effLat, "power[T1] ~ Y1")) * 1.1
yl <- range(0, eff(effObs, "Y2 ~ power[T1]"), eff(effLat, "Y2 ~ power[T1]")) * 1.1
shortName <- function(i) sub("_role$", "", i)
```

```{r, fig.height = 6.5}
plot(effObs, type = "effects", labels = shortName, xlim = xl, ylim = yl,
     title = "No reliability adjustment")
```

```{r, fig.height = 6.5}
plot(effLat, type = "effects", labels = shortName, xlim = xl, ylim = yl,
     title = "Modeled reliability of .70 for each adjective")
```

A point's fill shows its change effect (filled if p < .05, white otherwise),
and its outline shows its selection effect (black if p < .05, light gray
otherwise). So filled black points have both effects significant, and filled
gray points a significant change effect without selection. Points in the upper-right and
lower-left quadrants have selection and change effects of the same sign: the
relationship is *corresponsive* (Roberts, Caspi, & Moffitt, 2003). Members
high on the trait have more power, and power moves them further in the same
direction. Without adjustment, 16 adjectives show a significant change
effect, and the points rise along a diagonal: agentic and extraverted
adjectives such as *powerful*, *persuasive*, *dominant* and *extraverted* sit
at the upper right, and *shy*, *quiet* and *introverted* at the lower left.
Every adjective with a sizeable effect in both directions has them in the same
direction. A few adjectives, such as *relaxed*, *careless* and *unemotional*,
sit near the vertical axis: power predicts change in them without having
selected on them, as with `Relaxed_role` above.

At a modeled reliability of .70, the selection effects spread out, as
adjusted concurrent associations should, but only six change effects remain
significant: *powerful*, *shy* and *trustful*, plus *careless*, *relaxed* and
*unemotional*. Much of the evidence that power changes agentic role identities
rests on how completely Time 1 standing is controlled. *Powerful* and *shy*
are the robust cases.

Note that selection here is the *concurrent* association of a trait with
power. The non-correspondence Wood and Harms described contrasts Path B with
Path A, the traits that predict *gaining* power, which the next section takes
up.

## Do traits change power? (Path A)

Now power is the item, and a Time 1 trait is a control. Wood and Harms focused
on the overall communion and agency of people's trait profiles. A small
helper pulls out a control's cross-lagged coefficient, `Y2 ~ C`, from the
fitted models:

```{r}
crossPath <- function(x, from) {
  cf <- x$coefficients
  cf <- cf[cf$op == "~" & cf$rhs == from & cf$lhs == paste0(cf$item, "[T2]"),
           c("item", "est", "se", "pvalue")]
  cf[c("est", "se", "pvalue")] <- lapply(cf[c("est", "se", "pvalue")], round, 3)
  rownames(cf) <- NULL
  cf
}
pAc <- stabilityPaths(d, items = "power", controls = c("communion[T1]", ctrl))
pAa <- stabilityPaths(d, items = "power", controls = c("agency[T1]", ctrl))
print(pAc, pool = list(organization = orgs))
rbind(crossPath(pAc, "communion[T1]"), crossPath(pAa, "agency[T1]"))
```

Communion predicts gains in power (Wood and Harms reported a partial
correlation of .20). Agency, despite being the stronger *concurrent* correlate
of power, predicts gains more weakly (.08 in the paper). This is the
non-corresponsive pattern: the traits power seems to foster are not the ones
that earn it.

The communion route carries only a small share of power's stability. Path A is
clearly positive, but communion and power are only modestly related at Time 1,
and the pathway is the product of the two links. A trait can predict *change*
in power without accounting for much of power's *continuity*.

Tenure is also informative. Its pathway is negative: longer-serving members
had more power at Time 1 but tended to lose some of it over the year, as
seniors made way for the next cohort.

Profile communion and agency are highly correlated (about .8), because many
adjectives are desirable in both senses. Entering both at once asks which one
matters once the other is held constant:

```{r}
pAboth <- stabilityPaths(d, items = "power",
                         controls = c("communion[T1]", "agency[T1]", ctrl))
rbind(crossPath(pAboth, "communion[T1]"), crossPath(pAboth, "agency[T1]"))
```

Communion carries the effect. Holding communion constant, agency, if
anything, predicts *losing* power. This matches the lower half of Wood and
Harms's Table 1. Because the two profiles overlap so heavily, these
coefficients have larger standard errors than either profile alone.

## How much do organizations B and D matter?

Two organizations have a problem with how IDs link records. In organization D,
the IDs on Time 1 surveys don't appear to match the IDs of the same people on
the Time 1 influence roster. In organization B, roster IDs don't appear to
refer to the same people at the two waves. The evidence is summarized in
`?powerTraits`; `people$linkageIssue` flags both organizations.

Both were included in the published analyses, and including them reproduces
the published results closely. Refitting without them shows how much the
conclusions depend on them. Without B and D, their dummies are all zero, so
they leave the controls:

```{r}
dBD <- d[!d$linkageIssue, ]
ctrlBD <- setdiff(ctrl, c("orgB", "orgD"))
keyPaths <- function(dat, ctrl) {
  a_c <- stabilityPaths(dat, items = "power", controls = c("communion[T1]", ctrl))
  a_a <- stabilityPaths(dat, items = "power", controls = c("agency[T1]", ctrl))
  b   <- stabilityPaths(dat, items = c("Powerful_role", "Shy_role"),
                        X = "power[T1]", controls = ctrl)
  c(`A: communion -> power` = crossPath(a_c, "communion[T1]")$est,
    `A: agency -> power`    = crossPath(a_a, "agency[T1]")$est,
    setNames(summary(b)$structural[["Y2 ~ power[T1]"]],
             paste("B: power ->", c("Powerful_role", "Shy_role"))))
}
round(cbind(allOrganizations = keyPaths(d, ctrl),
            withoutBandD     = keyPaths(dBD, ctrlBD)), 3)
```

Without B and D, Path B for the role identities holds up. Path A for
communion is weaker and less certain in the smaller sample. Whether B and D
should be kept depends on whether their ID links can be verified; until then,
it is worth reporting results both ways.

## Interpreting shares, and some cautions

- **Shares are shares of stability**, not of variance in Time 2 scores. A
  small mediated share can still reflect a large effect of the experience
  when selection is weak, as `Relaxed_role` shows.
- **Timing.** The mediated pathway treats the experience as coming between
  the waves. Here power was measured at Time 1, alongside the Time 1
  self-ratings, so the direction of the selection link is assumed rather than
  observed (Maxwell & Cole, 2007). The same applies whenever an experience is
  measured at Time 2 or spans the whole interval.
- **Unmeasured confounding.** Controls address only the confounders you
  measure. A stable variable left out of `controls` inflates both the
  socialization path and the residual path.
- **Totals near zero** make shares unstable. Read shares alongside the
  estimates.
- **Binary experiences**, such as whether someone held an office, should not
  be given a reliability: a record of whether something happened isn't one
  answer sampled from a range of answers a person might give.

## Toward cross-lagged models

Paths A and B are the two cross-lagged paths of a two-wave cross-lagged panel
model, and the two decompositions above are its two halves. Here each
direction was fitted separately, one trait at a time. A dedicated
cross-lagged model would fit both directions together. It would give both
decompositions from one set of estimates, and it could also split the Time 2
association between a trait and power into what carries over from Time 1,
what the cross-lagged paths produce, and change that the two share.

## References

Chmielewski, M., & Watson, D. (2009). What is being assessed and why it
matters: The impact of transient error on trait research. *Journal of
Personality and Social Psychology, 97*(1), 186--202.

Enders, C. K. (2010). *Applied missing data analysis*. Guilford Press.

Maxwell, S. E., & Cole, D. A. (2007). Bias in cross-sectional analyses of
longitudinal mediation. *Psychological Methods, 12*(1), 23--44.

McCrae, R. R., Kurtz, J. E., Yamagata, S., & Terracciano, A. (2011).
Internal consistency, retest reliability, and their implications for
personality scale validity. *Personality and Social Psychology Review,
15*(1), 28--50.

Roberts, B. W., Caspi, A., & Moffitt, T. E. (2003). Work experiences and
personality development in young adulthood. *Journal of Personality and Social
Psychology, 84*(3), 582--593.

Rubin, D. B. (1976). Inference and missing data. *Biometrika, 63*(3), 581--592.

Schafer, J. L., & Graham, J. W. (2002). Missing data: Our view of the state of
the art. *Psychological Methods, 7*(2), 147--177.

Watson, D. (2004). Stability versus change, dependability versus error:
Issues in the assessment of personality over time. *Journal of Research in
Personality, 38*(4), 319--350.

Wood, D., & Harms, P. D. (2017). Evidence of non-corresponsive causal
relationships between personality traits and social power over time.
*Personality and Social Psychology Bulletin, 43*(1), 33--45.

Wood, D., Lowman, G. H., Armstrong, B. F., III, & Harms, P. D. (2023). Using
retest-adjusted correlations as indicators of the semantic similarity of
items. *Journal of Personality and Social Psychology, 125*(2), 437--454.
https://doi.org/10.1037/pspp0000441

Wood, D., & Roberts, B. W. (2006). Cross-sectional and longitudinal tests of
the Personality and Role Identity Structural Model (PRISM). *Journal of
Personality, 74*(3), 779--809.
