---
title: "Cross-lagged effects: traits and social power"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-lagged effects: traits and social power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, out.width = "100%",
                      dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png",
                      dpi = 96, fig.retina = 2)
```

```{r setup}
library(fancyr)
```

## The question

People's traits and their experiences can shape each other. A trait can lead
people into certain experiences (*selection*), and experiences can change
traits (*change*, or socialization). Roberts, Caspi and Moffitt (2003)
proposed a *corresponsive principle*: the traits that lead people into an
experience tend to be the ones the experience then reinforces.

When a trait and an experience are both measured at two times, a two-wave
**cross-lagged panel model** estimates both directions at once. For each
item (a trait), `crossLagPaths()` fits:

| effect | path | question |
|---|---|---|
| **selection** | `X2 ~ Y1` | does the Time 1 trait predict the experience at Time 2, beyond the experience at Time 1? |
| **change** | `Y2 ~ X1` | does the Time 1 experience predict the trait at Time 2, beyond the trait at Time 1? |
| **stability** | `Y2 ~ Y1`, `X2 ~ X1` | how well does each variable predict itself? |
| **T1 association** | `Y1 ~~ X1` | how are the two related at Time 1? |
| **co-change** | `Y2 ~~ X2` | what do the Time 2 scores still share once everything at Time 1 is accounted for? |

Corresponsiveness means that selection and change effects have the same sign.

## The study

This vignette works through the study behind Wood and Harms (2017). Members of
seven fraternities and sororities were surveyed twice, a year or more apart.
At each wave they described themselves on 59 trait adjectives, twice over:
once as they see themselves **in general** (their *general identity*), and
once as they see themselves **as a member of their organization** (their
*role identity*). They also rated how much influence every member they knew
had within the organization. A person's **social power** is the average of
the influence ratings they received, typically from 20 to 50 other members,
so it is a very dependable estimate.

Wood and Harms asked two questions about traits and power over time:

| path | question | in the cross-lagged model |
|---|---|---|
| **A** | Do traits predict *changes* in power? | selection: `power[T2] ~ Y1` |
| **B** | Does power predict *changes* in traits? | change: `Y2 ~ power[T1]` |

The answer was that the two directions differ: the relationship is
*non-corresponsive*. Communal traits predicted gains in power (Path A), while
power predicted gains in agentic role identities, such as seeing oneself as
powerful and not shy (Path B).

### Why a cross-lagged model?

Here the experience, power, was measured at the same occasions as the
traits: each wave's power comes from the influence ratings members made in
that wave's survey. The cross-lagged model fits that design. Each path runs
from one occasion to the next, and each variable's own earlier score is
controlled.

`stabilityPaths()` answers a different question: how much of an item's
stability is carried by an experience that happens *between* the two
measurements of the item, such as a job taken up or an office held during
the interval. It treats the experience as coming after Time 1 and before
Time 2. With power measured at Time 1, alongside the Time 1 self-ratings,
that ordering would be assumed rather than observed, so that model is not
used here.

When each variable is measured matters for both models. The next sections
look at when these people were measured.

## The data

`powerTraits` is a list. `T1` and `T2` hold each wave's data, one row per
person, including the survey `date`. `people` holds fixed information about
each person. `items` gives the agency and communion content of each
adjective, as rated by judges.

```{r}
str(powerTraits, max.level = 1)
powerTraits$T1[1:3, 1:8]
head(powerTraits$people, 3)
```

Each wave has `power`, the 59 general-identity adjectives (`Kind`,
`Dominant`, ...), the same adjectives as role identities (`Kind_role`, ...),
and four profile scores: `communion`, `agency`, `communion_role` and
`agency_role`. A profile score is the within-person correlation between a
person's ratings on the 59 adjectives and the adjectives' communion (or
agency) content. It is high for someone who endorses communal adjectives and
rejects uncommunal ones. See `?powerTraits` for details.

`stabilityData()` puts the pieces together. Items measured at both waves,
named in `commonItems`, get `[T1]` and `[T2]` suffixes. Other columns found in
only one wave are dropped by default, with a message; `keep = "T1"` or
`keep = "all"` retain them. Further files, such as `people`, are joined on the
ID. Given the name of a date column, `stabilityData()` also computes each
person's `interval_days`.

```{r}
pt <- powerTraits
adj <- pt$items$item
d <- stabilityData(pt$T1, pt$T2, pt$people, date = "date",
                   commonItems = c("power", "communion", "agency",
                                   "communion_role", "agency_role",
                                   adj, paste0(adj, "_role")))
```

### When were people measured?

A cross-lagged effect describes what one variable predicts about another *a
given time later*. Its size depends on that lag: over a short lag, a
variable has little time to change, and over a very long one, the effect of
Time 1 may have faded. So the lag belongs in any description of the results.

```{r}
round(t(sapply(split(d$interval_days, d$org), function(v)
  c(people = sum(!is.na(v)), median = median(v, na.rm = TRUE),
    min = min(v, na.rm = TRUE), max = max(v, na.rm = TRUE)))))
```

The organizations fall into two groups. A, C, E and F were first surveyed in
spring 2004 and again about 12 months later. B, D and G were first surveyed
in fall 2004 and again in spring 2006, about 16 to 17 months later. Within an
organization, each wave's surveys were completed within about two months, so
the interval varies much less within organizations than between the two
groups.

The estimates below therefore average over lags of about one year and about
a year and a half. The organization dummies (below) absorb differences in
average levels between organizations. They don't allow the effects
themselves to differ with the lag. Note also that both organizations with ID
linkage problems (B and D, see *How much do organizations B and D matter?*)
are in the longer-lag group.

### Who is in the sample?

About half the Time 1 participants never answered the Time 2 survey, mostly
because seniors graduated. By default, `stabilityData()` **keeps** them:

```{r}
has <- function(v) !is.na(d[[v]])
c(everyone = nrow(d),
  selfRatingsBothWaves = sum(has("communion[T1]") & has("communion[T2]")),
  traitT1_powerT2      = sum(has("communion[T1]") & has("power[T2]")),
  powerT1_traitT2      = sum(has("power[T1]") & has("communion[T2]")))
```

That's deliberate. `crossLagPaths()` estimates its models by *full
information maximum likelihood* (FIML), which uses every value each person
has. Someone seen only at Time 1 still tells us about the Time 1 distribution
and about how Time 1 scores relate to power and the controls. Here that
matters doubly, because power comes from other members' ratings: many people
who skipped the Time 2 survey still have Time 2 power. As the counts show,
more people link a Time 1 trait to Time 2 power (the data behind selection)
than link Time 1 power to a Time 2 trait (the data behind change).

Dropping incomplete cases (listwise deletion, `join = "inner"`) is unbiased
only if dropout is *missing completely at random*: unrelated to anything,
measured or not. FIML only needs dropout to be *missing at random*: it may
depend on observed things like Time 1 scores, as long as it doesn't further
depend on the unobserved Time 2 score itself (Rubin, 1976; Schafer & Graham,
2002; Enders, 2010). The *Missing data* section of `?stabilityData` has more
detail and references.

### Controls: organization and tenure

Organizations differ in their average power and in how their members see
themselves, and Wood and Harms controlled for organization in every analysis
with dummy codes. We do the same: six indicator variables, with organization A
as the reference, join the controls, along with `tenure`, the years a person
had been a member at Time 1.

```{r}
orgs <- paste0("org", LETTERS[2:7])
d[orgs] <- lapply(LETTERS[2:7], function(o) as.numeric(d$org == o))
ctrl <- c("tenure", orgs)
```

Organizations are single-sex, so the dummies also control for gender. The
controls predict both Time 2 variables. Each dummy also has its own pathway
in the stability decompositions described below. To keep the printed tables
readable, `print()` and `plot()` can combine them with
`pool = list(organization = orgs)`. This affects the display only.

## Cross-lagged effects of the trait profiles

Wood and Harms focused on the overall communion and agency of people's trait
profiles. `crossLagPaths()` fits one model per profile, with power as `X`:

```{r}
profiles <- c("communion", "agency", "communion_role", "agency_role")
clP <- crossLagPaths(d, items = profiles, X = "power", controls = ctrl)
print(clP, pool = list(organization = orgs))
```

Communion predicts gains in power: the selection effect is .18 (Wood and
Harms reported a partial correlation of .20), and .15 for the communion of
role identities. Agency, although it is the stronger *concurrent* correlate
of power (the T1 association), predicts gains more weakly and not
significantly (.10; the paper reported .08). Power predicts no significant
change in any profile. This is the non-corresponsive pattern: the traits
power seems to foster are not the ones that earn it.

Each fit also decomposes *both* stabilities into additive parts, which sum to
the total (in this standardized metric, the retest correlation). The item's
stability gains a cross-lagged pathway through `power[T1]` (the T1
association times the change effect). Power's stability gains one through
`Y1` (the T1 association times the selection effect). Both are small here,
at 7% or less of either total. A trait can predict *change* in power without
accounting for much of power's *continuity*. Tenure matters more to power's
stability: longer-serving members had more power at Time 1 but tended to lose
some of it over the interval, as seniors made way for the next cohort.

The co-change is left undecomposed. It collects whatever the Time 2 scores
share beyond what Time 1 predicts. For the agency of role identities it is
.30: members whose role identities became more agentic also gained power
over the same interval. The model can't say whether one change caused the
other, or whether something else moved both.

`summary(clP)` gives standard errors, confidence intervals and p-values for
every effect and pathway, and `plot(clP, type = "bars")` draws the two
decompositions side by side.

### Communion and agency together

Profile communion and agency are highly correlated (about .8), because many
adjectives are desirable in both senses. Adding the other profile's Time 1
score to the controls asks which one matters once the other is held
constant:

```{r}
jc <- crossLagPaths(d, items = "communion", X = "power",
                    controls = c("agency[T1]", ctrl))
ja <- crossLagPaths(d, items = "agency", X = "power",
                    controls = c("communion[T1]", ctrl))
sel <- function(x) x$effects[x$effects$effect == "selection",
                             c("item", "path", "est", "se", "pvalue")]
rbind(sel(jc), sel(ja))
```

Communion carries the effect (.31). Holding communion constant, agency, if
anything, predicts *losing* power (−.16). This matches the lower half of Wood
and Harms's Table 1. Because the two profiles overlap so heavily, these
coefficients have larger standard errors than either profile alone.

## Adjusting for retest reliability

A single answer reflects the particular occasion as well as where the person
usually stands. That matters for both cross-lagged paths. The change effect
asks whether power predicts Time 2 self-ratings *over and above* Time 1
self-ratings. A single Time 1 answer only partly controls for the person's
expected score around Time 1. The rest of that expected score can be
credited to power, whenever power is related to it. Wood and Harms raised
exactly this concern about their Path B estimates. The selection effect is
affected in the other direction: occasion-specific influences on the Time 1
self-rating dilute its relation to later power.

If you know a variable's reliability, ideally a retest correlation over a few
weeks (see *Choosing a reliability* below), supply it. The variable is then
modelled as a latent variable measured with that reliability. Any variable
can be adjusted: the items, `X` (by its base name for both waves, as in
`power = .93`, or one wave at a time, as in `"power[T1]" = .93`), and the
controls. Anything not named is analysed as observed.

## Choosing a reliability

An item given a reliability stands for the person's *expected score* on it
over some interval *t*: roughly, the average of the answers they would give if
asked repeatedly across that span. Nothing assumes the expected score is
fixed. It can change between the waves, which is exactly what the model
studies. The reliability to supply is the share of variance in a single
answer that is shared with the expected score over *t*.

A retest correlation over an interval of about *t* estimates this directly
(Wood et al., 2023). Whatever contributes to both answers, meaning whatever
persists over *t*, is part of the expected score. Influences that come and go
within *t*, such as mood, the day's circumstances, or remembering an answer
given a moment ago, still shape each answer systematically. They just aren't
part of the expected score over *t*. As *t* stretches from an hour to a day,
a week, a month or a year, more of what lies behind an answer has the chance
to change, and retest correlations fall (Watson, 2004).

So choosing *t* decides what the adjusted results describe:

- **Short *t*.** With a same-session retest, or one a few days or a week
  apart, nearly everything behind an answer persists. The reliability is high
  and the adjustment small, and the adjusted results still describe scores
  tied to the particular occasion.
- **Long *t*.** As *t* approaches the gap between the waves, the change the
  model is meant to study gets folded into the reliability, and the
  adjustment does too much. If you use the Time 1–Time 2 retest correlation
  itself, the adjusted stability is 1 by construction.
- **In between.** Long enough that what is specific to an occasion has turned
  over, but short relative to the interval being studied. For waves a year
  apart, retest correlations over two to four weeks are a reasonable target.

Internal consistency (alpha) fits the same logic, because its items are
answered at different moments too, typically about a minute apart. Alpha is
often called an underestimate of reliability. What it underestimates is the
retest reliability of those items *over that short interval*. If two items
answered about a minute apart give an alpha of .80, each item's own retest
correlation over a minute (averaged, say, as a geometric mean) should exceed
.80. Over longer intervals such as two weeks, observed retest correlations are
regularly *lower* than alpha (McCrae et al., 2011; Chmielewski & Watson,
2009). So alpha isn't a suitable value for waves months or years apart, and a
single item has no alpha anyway.

The reliability of power is a different kind of number. Because power
averages many raters, the reliability of each person's average is about .93
(the mean across organizations, estimated from how much raters agreed). It
describes how much a person's average would change with a different set of
raters, not with a different occasion. Over a few weeks, few members' power
is likely to shift much, so it serves as a reasonable approximation here, but
the two quantities are not the same.

Suitable retest values often don't exist for the items you have. Published
retest correlations for similar items or measures can suggest a range, and
the sensitivity analysis below shows whether the conclusions hold across it.

## How much does the modeled reliability matter?

Standard errors treat a supplied reliability as known, but it's an estimate
too, and conclusions can depend on it. `reliabilitySensitivity()` refits the
analysis across a grid of reliabilities for the items, keeping the rest as
they were (here power is observed, as in the paper), and records the effects
with their standard errors at each value. A value of 1 is the analysis
without adjusting the items.

```{r, fig.width = 8, fig.height = 4.5}
sens <- reliabilitySensitivity(clP, rel = c(.7, .85, 1))
sens
plot(sens)
```

Filled points have p < .05, and bars are 95% confidence intervals. Modeling
the profiles as less reliable strengthens communion's selection effect a
little (from .18 to .22 at a reliability of .70), and it stays significant.
The change effects shrink to about zero. Both halves of the non-corresponsive
pattern hold up under adjustment.

## Selection and change effects across all adjectives

Profiles summarize many adjectives at once. To see the adjectives
individually, `plot()` of a cross-lagged fit places each item by its
selection effect (horizontal) and change effect (vertical). Here are all 59
role identities, first without adjustment, then modeling each adjective's
reliability as .70 and power's as .93:

```{r}
roleAll <- paste0(adj, "_role")
clObs <- crossLagPaths(d, items = roleAll, X = "power", controls = ctrl)
clLat <- crossLagPaths(d, items = roleAll, X = "power", controls = ctrl,
                       reliability = c(setNames(rep(.7, length(roleAll)), roleAll),
                                       power = .93))
lim <- max(abs(unlist(lapply(list(clObs, clLat), function(x)
  x$effects$est[x$effects$effect %in% c("selection", "change")])))) * 1.08
shortName <- function(i) sub("_role$", "", i)
```

The same `xlim` and `ylim` put both plots on the same scale:

```{r, fig.height = 7.5}
plot(clObs, labels = shortName,
     xlim = c(-lim, lim), ylim = c(-lim, lim), title = "No reliability adjustment")
```

```{r, fig.height = 7.5}
plot(clLat, labels = shortName,
     xlim = c(-lim, lim), ylim = c(-lim, lim),
     title = "Modeled reliability of .70 for each adjective")
```

Both axes share one range, so selection and change effects of the same size
sit the same distance from zero. A point's fill shows which effects reached
p < .05: blue for change only, orange for selection only, black for both,
white for neither.

The shading shows where significance is possible. An estimate is significant
when it is more than 1.96 standard errors from zero, so every adjective has
its own critical value. Inside the darker band, an estimate would not be
significant for *any* adjective. In the lighter band, it would be for some
adjectives but not others. Beyond both, it would be significant for all of
them. Thin gray lines mark the band edges, so a point near zero on one axis
can still be read against the critical values of the other.

The bands are narrower for selection than for change. Selection effects are
estimated more precisely because more people contribute to them: Time 2
power is known for many members who skipped the Time 2 survey, while change
needs a Time 2 self-rating. So a selection effect of a given size is more
easily detected than a change effect of the same size, and a null selection
effect here is more informative than a null change effect.

Each plot's bands come from its own analysis, and they are wider under the
reliability adjustment: roughly ±.13 to ±.18 for selection and ±.18 to ±.24
for change, against ±.10 to ±.13 and ±.13 to ±.16 without it. The adjustment
adds no information about anyone's scores. It rescales the relationships to
what they would be among expected scores, and it rescales their sampling
error along with them. The standard errors grow by about 30%, more than the
estimates do. The adjusted Time 1 trait also overlaps more with Time 1 power,
so each cross-lagged path rests on less information that is unique to it. An
effect therefore has to be larger to reach significance once reliability is
modeled. That is why fewer change effects are significant at .70, even though
most estimates move only a little.

Points in the upper-right and lower-left quadrants would be *corresponsive*
(Roberts, Caspi, & Moffitt, 2003): the trait would earn power, and power
would push it further in the same direction. That isn't what the data show.
Without adjustment, 19 adjectives have a significant change effect. Power
raises agentic and extraverted role identities, such as *powerful*,
*persuasive*, *verbal*, *dominant* and *extraverted*. It lowers *shy*,
*quiet*, *introverted* and *reserved*, and also *relaxed*, *careless* and
*unemotional*. Yet none of these adjectives predicts *gaining* power: they sit
inside the selection bands. Only four have significant selection effects:
*sympathetic* (positive), and *cold*, *temperamental* and *moody*
(negative). The traits that earn power are communal and even-tempered, and
the traits power fosters are agentic. Across the 59 adjectives, selection
and change effects correlate only .10.

At a modeled reliability of .70, ten change effects remain significant, led
by *shy*, *relaxed*, *powerful* and *careless*. Only *cold* and
*temperamental* still predict (losing) power.

### Role and general identities

Role experiences should affect role identities first (Wood & Roberts, 2006).
The same adjectives rated as general identities show whether power's effects
reach how people see themselves in general:

```{r}
someAdj <- c("Powerful", "Persuasive", "Assertive", "Shy", "Relaxed")
chg <- function(x) round(x$effects$est[x$effects$effect == "change"], 3)
clRole <- crossLagPaths(d, items = paste0(someAdj, "_role"), X = "power",
                        controls = ctrl)
clGen  <- crossLagPaths(d, items = someAdj, X = "power", controls = ctrl)
data.frame(adjective = someAdj, role = chg(clRole), general = chg(clGen))
```

Except for *persuasive* (.18, p < .05), power's change effects on how people
see themselves in general are much weaker than on their role identities, and
none of the others is significant.

## How much do organizations B and D matter?

Two organizations have a problem with how IDs link records. In organization D,
the IDs on Time 1 surveys don't appear to match the IDs of the same people on
the Time 1 influence roster. In organization B, roster IDs don't appear to
refer to the same people at the two waves. The evidence is summarized in
`?powerTraits`; `people$linkageIssue` flags both organizations.

Both were included in the published analyses, and including them reproduces
the published results closely. Refitting without them shows how much the
conclusions depend on them. Without B and D, their dummies are all zero, so
they leave the controls:

```{r}
dBD <- d[!d$linkageIssue, ]
ctrlBD <- setdiff(ctrl, c("orgB", "orgD"))
keyEffects <- function(dat, ctrl) {
  f <- crossLagPaths(dat, items = c("communion", "agency", "Powerful_role", "Shy_role"),
                     X = "power", controls = ctrl)
  e <- f$effects[f$effects$effect %in% c("selection", "change"), ]
  setNames(e$est, paste(e$item, e$effect))
}
round(cbind(allOrganizations = keyEffects(d, ctrl),
            withoutBandD     = keyEffects(dBD, ctrlBD)), 3)
```

Without B and D, power's change effects on the role identities hold up. The
selection effect of communion is weaker (.13), in a smaller sample. The change
effects of power on the two profiles are larger (.14 and .12), so the
conclusion that power leaves overall communion and agency unchanged is less
secure without these organizations. Dropping them also removes two of the
three longer-lag organizations, so the comparison mixes the linkage problem
with the lag. Whether B and D should be kept depends on whether their ID
links can be verified; until then, it is worth reporting results both ways.

## Cautions

- **The lag.** Cross-lagged effects are specific to the interval between
  measurements. Here that is about a year for some organizations and about a
  year and a half for others. Effects over a month or five years could be
  quite different.
- **Stable differences between people.** The two-wave cross-lagged model
  treats each person's Time 1 score as the whole of what carries forward. If
  people also differ in stable ways that affect both variables, the
  cross-lagged paths can partly reflect those differences rather than
  within-person change (Hamaker, Kuiper, & Grasman, 2015). Separating the two
  needs three or more waves.
- **Unmeasured confounding.** Controls address only the confounders you
  measure. A Time 1 variable left out of `controls` that affects both
  Time 2 variables can masquerade as a cross-lagged effect.
- **Co-change is not an effect.** A large co-change says the two variables
  moved together during the interval. It doesn't say which moved which, or
  whether a third variable moved both.
- **Shares are shares of stability**, not of variance in Time 2 scores, and
  totals near zero make shares unstable. Read shares alongside the estimates.
- **Binary variables**, such as whether someone held an office, should not
  be given a reliability: a record of whether something happened isn't one
  answer sampled from a range of answers a person might give.

## References

Chmielewski, M., & Watson, D. (2009). What is being assessed and why it
matters: The impact of transient error on trait research. *Journal of
Personality and Social Psychology, 97*(1), 186--202.

Enders, C. K. (2010). *Applied missing data analysis*. Guilford Press.

Hamaker, E. L., Kuiper, R. M., & Grasman, R. P. P. P. (2015). A critique of
the cross-lagged panel model. *Psychological Methods, 20*(1), 102--116.

McCrae, R. R., Kurtz, J. E., Yamagata, S., & Terracciano, A. (2011).
Internal consistency, retest reliability, and their implications for
personality scale validity. *Personality and Social Psychology Review,
15*(1), 28--50.

Roberts, B. W., Caspi, A., & Moffitt, T. E. (2003). Work experiences and
personality development in young adulthood. *Journal of Personality and Social
Psychology, 84*(3), 582--593.

Rubin, D. B. (1976). Inference and missing data. *Biometrika, 63*(3), 581--592.

Schafer, J. L., & Graham, J. W. (2002). Missing data: Our view of the state of
the art. *Psychological Methods, 7*(2), 147--177.

Watson, D. (2004). Stability versus change, dependability versus error:
Issues in the assessment of personality over time. *Journal of Research in
Personality, 38*(4), 319--350.

Wood, D., & Harms, P. D. (2017). Evidence of non-corresponsive causal
relationships between personality traits and social power over time.
*Personality and Social Psychology Bulletin, 43*(1), 33--45.

Wood, D., Lowman, G. H., Armstrong, B. F., III, & Harms, P. D. (2023). Using
retest-adjusted correlations as indicators of the semantic similarity of
items. *Journal of Personality and Social Psychology, 125*(2), 437--454.
https://doi.org/10.1037/pspp0000441

Wood, D., & Roberts, B. W. (2006). Cross-sectional and longitudinal tests of
the Personality and Role Identity Structural Model (PRISM). *Journal of
Personality, 74*(3), 779--809.
